---
title: "Model and Jags Code"
output: pdf_document
date: "2026-04-22"
---

## Goal and Project Plan

We want be able to predict the outcome of matches for tennis players who do not have much data. We also want to know if there is a considerable court surface effect.  

This will be accomplished with the following steps:
1) Fit mixed effect models with players who have played over 50 matches with and without surface effects and interaction effects
2) Compare those models with DIC
3) Use the posteriors of the winning model as priors in a model with players who have only 10-50 matches recorded. We are particularly interested in the player random effect parameters

## Models for Experienced players

$Y_{ijk}$ is the outcome of the $k^{th}$ match for the $i^{th}$ player on the $j^{th}$ court surface
$Y_ijk \sim \text{Bernoulli}(p_{ijk})$
$X$ is the rank difference in the $k^{th}$ match. $Opponent - Player$
$A_i$ is the random effect of the $i^{th}$ player  $A_i \sim N(0, \sigma_A^2)$
$\alpha_j$ is the fixed effect of the $j^{th}$ court surface
$\beta_0, \beta_1 \sim N(0, 10^2)$
$E_{ijk} \overset{iid} \sim N(0, \sigma^2)$

M1: Fixed surface effect with interaction
$$ \text{logit}(p_{ijk}) = A_i + \alpha_j + (A \alpha)_{ij} + \beta_0 + \beta_1 X{ijk} + E_{ijk}$$
M2: Additive model with fixed surface effect
$$ \text{logit}(p_{ijk}) = A_i + \alpha_j + \beta_0 + \beta_1 X{ijk} + E_{ijk}$$
M3: Random effects model with no consideration with surface effects
$$ \text{logit}(p_{ijk}) = A_i + \beta_0 + \beta_1 X{ijk} + E_{ijk}$$

## Run.jags vs rjags (Delete before final product)
*Delete this later.* This is the difference in code between run.jags and rjags. The as.mcmc.list converts the output into what rjags returns. We can extract DIC by the code below
```{r}
jags_fit = run.jags(model = model_string, data=data, monitor=params,
                   n.chains=4, burnin=10000, sample=20000, modules="glm", 
                   method="parallel", silent.jags = TRUE)

jags_fit

# Convert to rjags output for familiar plots
samples = as.mcmc.list(jags_fit)
par(mar = c(1, 1, 1, 1)) 
plot(samples)

DIC = extract(jags_fit, what = "dic")
```

## Jags code
### Libraries
```{r}
source("Data_cleaning.R")
```

```{r}
library(tidyverse)
library(runjags)
library(rjags)
library(lme4)
```


### Preliminary

```{r}
data = list(
  N = nrow(top_df), 
  Y = top_df$outcome,
  player = top_df$player,
  surface = top_df$surface, 
  X = top_df$rank_diff |> scale() |> as.numeric(),
  n_players = length(unique(top_df$player)), 
  n_surfaces = length(unique(top_df$surface))
)

# frequentist methods to see if we can decrease the vagueness of our priors
# jags_df$player_surface = factor( paste(jags_df$player, jags_df$surface) )
# jags_df$y = jags_df$outcome-1
# glmer(y~surface+rank_diff+(1|player) + (1|player_surface), data=jags_df, family=binomial)

model_string = "model{

  #Likelihood
  for(i in 1:N){
    Y[i] ~ dbern(p[i])
    logit(p[i]) <- beta0 + beta1*X[i] +
                    A[player[i]] +
                    alpha[surface[i]] + 
                    gamma[player[i], surface[i]]
  }
  
  for(i in 1:n_players){
  
    # Player effect
    A[i] ~ dnorm(0, tau_A)
    
    for(j in 1:n_surfaces){
    
    # Interactoin effect
      gamma[i, j] ~ dnorm(0, tau_inter)
    }
  }
  
  # Surface effect
  for(j in 1:n_surfaces){
    alpha[j] ~ dnorm(0, .25)     ## SD = 2
  }
  
  # Other priors
  beta0 ~ dnorm(0, .25)
  beta1 ~ dnorm(0, .25)
  
  sigma_A ~ dt(0., 100., 2)T(0., )  ## Half-T with scale=0.1 and 2 df
  sigma_inter ~ dt(0., 100., 2)T(0., )
  tau_A <- 1/sigma_A^2
  tau_inter <- 1/sigma_inter^2
}"
```

```{r}
interaction_fit = run.jags(model = model_string, data=data, 
                   monitor=c("beta0", "beta1", "tau_A", "tau_inter"),
                   n.chains=4, burnin=10000, sample=20000, modules="glm", 
                   method="parallel")

#, silent.jags = TRUE) 

```

