---
title: "Model and Jags Code"
output: pdf_document
date: "2026-04-22"
editor_options: 
  markdown: 
    wrap: 72
---

## Goal and Project Plan

We want be able to predict the outcome of matches for tennis players who
do not have much data. We also want to know if there is a considerable
court surface effect.

This will be accomplished with the following steps: 1) Fit mixed effect
models with players who have played over 50 matches with and without
surface effects and interaction effects 2) Compare those models with DIC
3) Use the posteriors of the winning model as priors in a model with
players who have only 10-50 matches recorded. We are particularly
interested in the player random effect parameters

## Models for Experienced players

$Y_{ijk}$ is the outcome of the $k^{th}$ match for the $i^{th}$ player
on the $j^{th}$ court surface $Y_{ijk} \sim \text{Bernoulli}(p_{ijk})$
$X$ is the rank difference in the $k^{th}$ match. $Opponent - Player$
$A_i$ is the random effect of the $i^{th}$ player
$A_i \sim N(0, \sigma_A^2)$ $\alpha_j$ is the fixed effect of the
$j^{th}$ court surface $\beta_0, \beta_1 \sim N(0, 10^2)$
$E_{ijk} \overset{iid} \sim N(0, \sigma^2)$

M1: Fixed surface effect with interaction
$$ \text{logit}(p_{ijk}) = A_i + \alpha_j + (A \alpha)_{ij} + \beta_0 + \beta_1 X{ijk} + E_{ijk}$$
M2: Additive model with fixed surface effect
$$ \text{logit}(p_{ijk}) = A_i + \alpha_j + \beta_0 + \beta_1 X{ijk} + E_{ijk}$$
M3: Random effects model with no consideration with surface effects
$$ \text{logit}(p_{ijk}) = A_i + \beta_0 + \beta_1 X{ijk} + E_{ijk}$$

## Run.jags vs rjags (Delete before final product)

*Delete this later.* This is the difference in code between run.jags and
rjags. The as.mcmc.list converts the output into what rjags returns. We
can extract DIC by the code below

```{r eval = F}
jags_fit = run.jags(model = model_string, data=data, monitor=params,
                   n.chains=4, burnin=10000, sample=20000, modules="glm", 
                   method="parallel", silent.jags = TRUE)

jags_fit

# Convert to rjags output for familiar plots
samples = as.mcmc.list(jags_fit)
par(mar = c(1, 1, 1, 1)) 
plot(samples)

DIC = extract(jags_fit, what = "dic")
```

## Jags code

### Libraries

```{r}
source("Data_cleaning.R")
```

```{r}
library(tidyverse)
library(runjags)
library(rjags)
library(lme4)
```

### Preliminary

Since our models are nested within each other, we'll conduct SSVS
(Stochastic Search Variable Selection) to find the best variables. We
can to this by enforcing hierarchy within our model string and adding a single ordinal indicator variable.

```{r}
data = list(
  N = nrow(top_df), 
  Y = top_df$outcome,
  player = top_df$player,
  surface = top_df$surface, 
  X = top_df$rank_diff |> scale() |> as.numeric(),
  n_players = length(unique(top_df$player)), 
  n_surfaces = length(unique(top_df$surface))
)

# frequentist methods to see if we can decrease the vagueness of our priors
# jags_df$player_surface = factor( paste(jags_df$player, jags_df$surface) )
# jags_df$y = jags_df$outcome-1
# glmer(y~surface+rank_diff+(1|player) + (1|player_surface), data=jags_df, family=binomial)

model_string = "model{

  #Likelihood
  for(i in 1:N){
    Y[i] ~ dbern(p[i])
    logit(p[i]) <- beta0 + beta1*X[i] +
                    A[player[i]] +
                    # delta=0: M3, delta=1: M2, delta=2: M1
                    step(delta - 0.5) * alpha[surface[i]] + 
                    step(delta - 1.5) * gamma[player[i], surface[i]]
  }
  
  for(i in 1:n_players){
  
    # Player effect
    A[i] ~ dnorm(0, tau_A)
    
    for(j in 1:n_surfaces){
    
    # Interactoin effect
      gamma[i, j] ~ dnorm(0, tau_inter)
    }
  }
  
  # Surface effect
  for(j in 1:n_surfaces){
    alpha[j] ~ dnorm(0, .25)     ## SD = 2
  }
  
  # SSVS indicators
  # Single ordinal indicator: 0=M3, 1=M2, 2=M1
  # Uniform prior over three models
  delta ~ dcat(p_model[])   
  p_model[1] <- 1/3         
  p_model[2] <- 1/3
  p_model[3] <- 1/3
  
  # Other priors
  beta0 ~ dnorm(0, .25)
  beta1 ~ dnorm(0, .25)
  
  sigma_A ~ dt(0., 1., 4)T(0., )  ## Half-T with scale=1 and 4 df, changed to prevent overflow
  sigma_inter ~ dt(0., 1., 4)T(0., )
  tau_A <- 1/sigma_A^2
  tau_inter <- 1/sigma_inter^2
}"
```

## JAGS model

```{r}
ssvs_fit = run.jags(model = model_string, data=data, 
                   monitor=c("beta0", "beta1", "delta", "sigma_A", "sigma_inter"),
                   n.chains=4, burnin=10000, sample=20000, modules="glm", 
                   method="parallel")

#, silent.jags = TRUE) 
# This took me about 15 minutes
```


### Posterior model probabilities
```{r}
samples_mat <- as.matrix(as.mcmc.list(ssvs_fit))

# Posterior probability of each model
p_m3 <- mean(samples_mat[, "delta"] == 1)  # dcat indexes from 1
p_m2 <- mean(samples_mat[, "delta"] == 2)
p_m1 <- mean(samples_mat[, "delta"] == 3)

cat("P(M3 - player only):      ", round(p_m3, 3), "\n")
cat("P(M2 - additive surface): ", round(p_m2, 3), "\n")
cat("P(M1 - interaction):      ", round(p_m1, 3), "\n")
```
Model 3 is completely ruled out, but the posterior probability is split on model 1
and 2. May have to use DIC as a tiebreaker.

### Convergence diagnostics check

```{r}
samples = as.mcmc.list(ssvs_fit)
gelman.diag(samples)
effectiveSize(samples)
```


### Trace plots 
(not including delta (categorical))
```{r}
par(mar = c(2, 2, 2, 2))
traceplot(samples[, c("beta0", "beta1", "sigma_A", "sigma_inter")])
```

## Where to Go Next:

1.  SSVS model (as above) → interpret which model structure
    is best, possibly supported with DIC evaluation.
    
2.  Winning model: refit with full parameter monitoring (including A[i],
    alpha[j]) — you need posterior samples of all parameters to
    construct priors.
    
3.  Prior extraction — summarize the posteriors of
    beta0, beta1, sigma_A, and alpha[j] (means and precisions) to pass
    as informative priors into the mid-player model 
    
4.  Mid-player model
    using mid_df: same structure as the winning model but with tighter
    priors from step 3, and monitor the player random effects A[i]
    closely since that's your main scientific interest
    
5.  Results/comparison: compare posterior distributions of player
    effects between the two groups, maybe a shrinkage plot showing how
    the mid-player estimates are pulled toward the experienced-player
    posteriors

