---
title: "Bayesian Analysis Final Project"
author: "Ryan Mersereau"
date: "2026-04-22"
output: pdf_document
---
Install needed pacakges
```{r, include = F}
library(rjags)
library(tidyverse)
library(coda)
library(ggplot2)
```

Specify years, read in data from Jeff Sackmans ATP dataset direct from github 
```{r}
years <- 2019:2023
base_url <- "https://raw.githubusercontent.com/JeffSackmann/tennis_atp/master/atp_matches_"

atp_raw <- map_dfr(years, function(yr) {
  url <- paste0(base_url, yr, ".csv")
  read_csv(url, show_col_types = FALSE)
})
```

Clean and filter data
```{r}
# Keep only the columns we need, filter to top players and 3 surfaces
atp_clean <- atp_raw |>
  select(winner_name, loser_name, surface, tourney_date, winner_rank, loser_rank) |>
  filter(surface %in% c("Clay", "Hard", "Grass")) |>
  drop_na()

# Keep only players with enough total matches (e.g., 50+ matches across all surfaces)
# This keeps the model manageable and removes journeymen
match_counts <- bind_rows(
  atp_clean |> count(player = winner_name),
  atp_clean |> count(player = loser_name)
) |>
  group_by(player) |>
  summarise(total = sum(n)) |>
  filter(total >= 50)

top_players <- match_counts$player

atp_filtered <- atp_clean |>
  filter(winner_name %in% top_players,
         loser_name  %in% top_players)

cat("Matches:", nrow(atp_filtered), "\n")
cat("Players:", n_distinct(c(atp_filtered$winner_name, atp_filtered$loser_name)), "\n")
```

Encode for JAGS: JAGS needs integer indeces instead of strings, need to map players and surfaces to numbers

# Additions
- Model for total matches greater than 50 and no filters (only prior information) 
- Ensure player has one match per year (don't want players to retire in the middle of our timeframe)
- Compare two or more models (random slopes, random effects (test for effect of court surface)), find lowest DIC among models

