---
title: "Bayesian Analysis Final Project"
author: "Ryan Mersereau"
date: "2026-04-22"
output: pdf_document
---
Install needed pacakges
```{r, include = F}
library(rjags)
library(tidyverse)
library(coda)
library(ggplot2)
```

Specify years, read in data from Jeff Sackmans ATP dataset direct from github 
```{r}
years <- 2019:2023
base_url <- "https://raw.githubusercontent.com/JeffSackmann/tennis_atp/master/atp_matches_"

atp_raw <- map_dfr(years, function(yr) {
  url <- paste0(base_url, yr, ".csv")
  read_csv(url, show_col_types = FALSE)
})
```

Clean and filter data
```{r}
# Keep only the columns we need, filter to top players and 3 surfaces
atp_clean <- atp_raw |>
  select(winner_name, loser_name, surface, tourney_date, winner_rank, loser_rank) |>
  filter(surface %in% c("Clay", "Hard", "Grass")) |>
  drop_na()

# Keep only players with enough total matches (e.g., 50+ matches across all surfaces)
# This keeps the model manageable and removes journeymen
match_counts <- bind_rows(
  atp_clean |> count(player = winner_name),
  atp_clean |> count(player = loser_name)
) |>
  group_by(player) |>
  summarise(total = sum(n)) |>
  filter(total >= 50)



top_players <- match_counts$player

atp_filtered <- atp_clean |>
  filter(winner_name %in% top_players,
         loser_name  %in% top_players)

cat("Matches:", nrow(atp_filtered), "\n")
cat("Players:", n_distinct(c(atp_filtered$winner_name, atp_filtered$loser_name)), "\n")
```

Encode for JAGS: JAGS needs integer indeces instead of strings, need to map players and surfaces to numbers

# Additions
- Model for total matches greater than 50 and no filters (only prior information) 
- Ensure player has one match per year (don't want players to retire in the middle of our timeframe)
- Compare two or more models (random slopes, random effects (test for effect of court surface)), find lowest DIC among models

```{r}
# Creating rank difference (loser - winner)
temp = atp_clean |>
  mutate(rank_diff = loser_rank - winner_rank, .keep = "unused")

# Pivoting longer to make every player have a row
long_df = temp |>
  pivot_longer(
    cols = c(winner_name, loser_name),
    names_to = "role",
    values_to = "player"
  ) |>
  mutate(outcome = if_else(role == "winner_name", 1, 0)) |>  # 1 if winner, 0 if loser
  mutate(rank_diff = if_else(outcome == 1, rank_diff, -rank_diff)) |> # Rank_diff now opponent - player
  select(player, outcome, surface, tourney_date, rank_diff) # Selecting only variables of interest
long_df

# Converting the character values to unique numeric integers for jags to use
jags_df = long_df |>
  mutate(across(c(player, outcome, surface), as.factor)) |> 
  mutate(across(c(player, outcome, surface), as.numeric)) |> 
  select(-tourney_date)

# Players who have played >= 50 games
top_players = jags_df |>
  group_by(player) |>
  summarize(count = n()) |>
  filter(count >= 50) |>
  select(player)

top_df = jags_df |> filter(top_players %in% player)

# Players who have played 10 <= games < 50
mid_players = jags_df |>mid_playerany_of()s = jags_df |>
  group_by(player) |>
  summarize(count = n()) |>
  filter(count >= 10 & count < 50) |>
  select(player)

mid_df = jags_df |> filter(mid_players %in% player)
```

