---
title: "Picking the Safe Underdog"
subtitle: "A Bayesian Mixed-Effects Approach to ATP Tennis Prediction"
author: "Ryan Mersereau and Naomi Novick"
institute: "ST540 Final Project"
format: pptx
editor: visual
---

## Motivation

- Casual tennis viewers struggle to pick a player to root for
- Big stars are recognizable, but most players are unfamiliar
- Choosing an underdog is appealing, but risky without context of data
- Goal: Build a simple, data-informed model for picking a "safe" underdog, designed for casual viewers

## Background

- Want a model designed for casual viewers, not experts
- Should ignore complex factors (play style, endurance, etc.)
- Uses only what you can easily see or look up:
    - Players
    - Court surface
    - ATP ranking

## Research Gap

-  Many tennis models require detailed data unavailable to casual fans
-  Simpler alternatives (regression, ELO, head-to-head) lack coverage for low-data players
-  None are designed specifically for players with limited match history
-  We can fill this gap with a Bayesian hierarchical model, borrowing strength across players

## Data

- ATP matches 2019-2023 available from Sackmann public GitHub repository
- Examine surface effect: Clay, Hard, and Grass
- Key predictor: Rank difference (opponent - player)
- "Top players": >= 50 matches
- "Mid players": 10-49 matches

## Model (check this notation)
- Outcome: Bernoulli(p), logit(p) = $\alpha$[surface] + $\beta$X + A[player] + B[player, surface]
    - A[i] ~ N(0, $\sigma^2_A$): Player random effect (skill)
    - B[i,j] ~ N(0, $\sigma^2_{Aa}$): Player surface interaction
- Top-player posteriors used as informative priors for mid-player model

## Results: Surface & Rank Effects
- SSVS determined surface effect significant
- DIC comparison between models: 
- Rank difference is a significant positive predictor of winning probability


## Results: Posterior Predictions (plot?)
```{r}
#library(ggplot2)
#ggplot(top_df1, aes(rank_diff, col="top")) + 
#  geom_density() + 
#  facet_grid(~surface)
```
- Bayesian credible intervals are wider for mid-players, more uncertainty
- Frequentist GLM estimates lack shrinkage and are more extreme at tails

## Summary (another plot?)


## Strengths & Limitations
- Shrinkage improves stability for low-data players
- Credible intervals communicate prediction uncertainty honestly
- Rank difference along is a noisy signal, ignores form, injuries, head-to-head
- Could expand time frame to capture larger player pool

## Only plot

```{r}
A = 1:10
plot(A)
```


## Gory Details {visibility "uncounted"}
- SSVS: `delta ~ dcat([1/3, 1/3, 1/3])`; P(M1)=0.49, P(M2)=0.51, P(M3)≈0
- Priors on σ: Half-T(0,1,4) truncated at 0; tau = 1/σ² ~ Gamma(a,b) for mid model
- Gamma hyperparameters fit via MLE on top-player posteriors: `optim(BFGS)`
- 4 chains, 10k burn-in, 20k samples; Gelman-Rubin R-hat < 1.05
- Surface encoding: 1=Clay, 2=Grass, 3=Hard (alphabetical factor levels in R)


## Script

*Script:*

*Imagine you turn on the TV and a game of tennis is playing. You don't necessarily follow tennis very closely, but you like watching the games and you want to know who you should root for. This is really easy for the big names like Nadal, Federer, Djokovic, Alcaraz, but what about the smaller names? What if you want to root for the underdog, but don't want to risk getting your hopes up for nothing?*

*My partner and I both casually enjoy tennis, and we thought this would be a fun algorithm to use next time we tune in to a match. We found this amazing repository from gitHub that has records of professional matches from 1968-2024. We have chose to hone in on data from the years 2019-2023 for completeness and recency.*

*Now there are a number of factors determining who will win a match of tennis. Handed-ness, endurance, play style, but remember, we are causal fans, we don't know any of this, nor do we know the intricate rock-paper-scissors mechanics of these factors. We are assuming that you can only see what is on TV and only have the motivation to google who the underdog is in a given scenario. So, we know the player, the court surface, and the men's professional tennis (ATP) ranking. What now?*
